---
title: "Cross Spectrum Computation"
author: "Andrew J Barbour"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross Spectrum Computation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<!-- bibliography: REFS.bib -->


Here I show how the modeling tools in **kitagawa** can be
used to study actual data.  Specifically, I will show
records of strain and pore-pressure from borehole
stations in the [Network of the Americas(https://www.unavco.org/nota/)^[previously known as the [Plate Boundary Observatory](https://www.unavco.org/)]
in a manner similar to the approaches taken in 
[Barbour (2015)](https://doi.org/10.1002/2015JB012201).

```{r, echo=TRUE}
library(kitagawa)
```

## Pore Pressure Changes from Teleseismic Strains

We first load the [psd](https://cran.r-project.org/package=psd) package, which includes a
suitable dataset for this example. In particular, we're 
interested in assessing the frequency-dependent 
relationship between pore pressure $p$ and
_areal_ strain $E_A$^[Relative changes in borehole
diameter, which can be related to volume strain in the rock] during the seismic
portion of the record.]

```{r, echo=TRUE}
library(psd)
data(Tohoku)
toh_orig <- with(subset(Tohoku, epoch=='seismic'), {
  cbind(
    scale(1e3*areal, scale=FALSE), # scale strain to nanostrain, remove mean
    scale(1e2*pressure.pore, scale=FALSE) # scale hPa to Pa, remove mean
  )
})
colnames(toh_orig) <- c('input','output')
toh.dat <- window(ts(toh_orig), 100, 2400)
```

Note how the records of this earthquake -- the 2011 $M_W 9$ Tohoku-Oki earthquake some thousands of
kilometers away -- are very nearly a mirror image of each other:
```{r, echo=FALSE, fig.show='hold', fig.width=7., fig.height=4.5}
library(RColorBrewer)
Set1 <- brewer.pal(8, 'Set1')
par(mar=c(3,3,0.2,0.2))
plot(toh.dat, yax.flip = TRUE, main="Strain and Pressure: 2011 M9 Tohoku")
```
The energy carried by the seismic wavetrain is focused predominately at long
periods and for the surface waves it is very nearly harmonic.
Zooming in on the early part of the Rayleigh wave:
```{r, echo=FALSE, fig.show='hold', fig.width=7., fig.height=4.5}
windat <- scale(window(toh.dat, 1400, 1600))
plot(windat[,'input'], lty=5, type='l', ylab='', main='Rescaled Input and Output')
lines(-windat[,'output'], col=2, lwd=1.5)
lines(as.vector(time(windat)), scale(apply(windat,1, function(x){x <- abs(x); atan2(x[1],x[2])}))/2, lty=1, col=4)
legend('topleft', c('Input strain','Output pressure','Internal angle'), col=c(1,2,4), lty=c(2,1,1), lwd=c(1,1.5,1))
```
we see a close mirroring of the input and the output, with the input occurring
before the output. The tangent of the angle between the signals 
during the periods of more purely harmonic strain 
is suggestive of a simple phase
lag. In other words, the pore pressure lags the strain.
These observations are consistent with the 
theory of linear poroelasticity, which predicts the following response in undrained conditions
(assuming an undrained Poisson's ratio of $1/3$):
$$
p \approx - \frac{4}{3} B \mu E_A
$$
where $B$ is the Skempton's coefficient and $\mu$ is the elastic
shear modulus of the fluid-saturated rock. 
All of this indicates that the pore pressure response can be modeled as a convolution of an input
signal (dynamic strain) and transfer function ($p = G \star E_A$). 

In 
this case the (scalar) proportionality implied by the timeseries is 
```{r, echo=TRUE}
m <- lm(output ~ input - 1, as.data.frame(toh.dat))
strain_scaling <- coef(m)
signif(strain_scaling, 3) # GPa/strain
```
but we will see how this is actually frequency dependent.
```{r, echo=FALSE, fig.show='hold', fig.width=4.5, fig.height=4.5}
IO <- as.matrix(toh.dat)
plot(IO[,1], IO[,2], 
     asp=1, col=NA, 
     main='Pressure-strain correlation',
     xlab="Input (strain)", 
     ylab="Output (pore pressure)")
grid()
points(IO[,1], IO[,2], pch=3)
abline(m, col=2)
```

## Cross-Spectrum Estimation

If the results of the spectral computation include the complex 
auto spectra and cross spectrum $[S_{11}, S_{12}, S_{22}]$, 
the _coherence_ spectrum $\gamma^2$ can be calculated by
$$
\gamma^2 = \frac{\left|S_{12}\right|^2}{S_{11} S_{22}},
$$
the _admittance_ spectrum  (or _gain_) $G$ can be calculated from
$$
G = \gamma \sqrt{S_{22} / S_{11}},
$$
and the phase spectrum $\Phi$ can be calculated from
$$
\Phi = \arg{S_{12}}
$$
As Priestley (1981) shows, the multitaper coherency spectrum ($\gamma$) can be described by an \texit{F} distribution:
$$
\frac{2 k \gamma}{(1-\gamma)} \sim F(2,4k)
$$
Hence, the probability that the absolute coherency is greater than $c$ is 
$$
P(|\gamma| \geq c, k) = (1 - c^2)^{k-1}
$$
```{r}
k <- 2*130 # number to start out with
gam <- seq(0.001, 1, by=0.001)
gamrat <- 2 * gam / (1 - gam)
Pgam <- pf(k*gamrat, 2, 4*k)
```
```{r, echo=FALSE, fig.show='hold', fig.width=5.5, fig.height=4.5}
k2 <- 100
Pgam2 <- pf(k2*gamrat, 2, 4*k2)
k3 <- 10
Pgam3 <- pf(k3*gamrat, 2, 4*k3)
x.g <- ((1 - gam)*gamrat/2)
plot(x.g, Pgam, type='l', 
     main=expression(F(2*","~4*k)), 
     xlab=expression(gamma), 
     ylab=expression(p(gamma,k)), log='x')
lines(x.g, Pgam2, lty=5)
lines(x.g, Pgam3, lty=2)
legend('bottomright', parse(text=c(sprintf("k==%s",c(k,k2,k3)))), lty=c(1,5,2))
```

The standard error in the admittance follows from the coherence spectrum:
$$
\sqrt{(1 - \gamma^2)/k}
$$

## Application to Tohoku record

First let's use [psd](https://cran.r-project.org/package=psd)
to estimate a cross spectrum between pressure and strain,
treating strain as the input to the system and
pressure as the output.
```{r, echo=TRUE}
#!order_matters
class(toh.dat)
toh_to_pspec <- toh.dat[,c('input','output')]
toh.cs <- psd::pspectrum(toh_to_pspec, ntap.init=k, verbose=FALSE)
```
Which gives the following results:
```{r, echo=TRUE}
class(toh.cs)
str(toh.cs)
```

We form the requisite quantities, which are included as output from `pspectrum`:
```{r, echo=TRUE, fig.show='hold', fig.width=5., fig.height=7}

f <- as.vector(toh.cs[['freq']]) # frequency
lf <- log10(f)
p <- 1/f # period

# coherence
Coh <- toh.cs[['coh']]
# wrapped phase (in radians)
Phi <- as.vector(toh.cs[['phase']])
suppressPackageStartupMessages(can.unwrap <- require(signal))
if (can.unwrap){
  # unwrap if possible
  Phi <- signal::unwrap(Phi)
}
# Admittance or Gain
G <- Mod(toh.cs[['transfer']])
G <- Coh * G
# Tapers
K <- toh.cs[['taper']] # 'tapers' object
k <- as.numeric(K)
# Uncertainty in the admittance
G.err <- sqrt((1 - Coh) / k)
```

<!-- with(toh.cs,{ -->
<!--   layout(matrix(1:3,nc=1)) -->
<!--   par(mar=c(3,3,1,1), mgp=c(2,0.5,0), tcl=0.5, oma=c(2,0,0,0)) -->
<!--   f <- freq -->
<!--   xlim <- c(1/200,0.5) -->
<!--   plot(f, Coh, type='l', xlim=xlim, log='x', ylim=c(0,1), ylab="Coherence") -->
<!--   plot(f, Phi, xlim=xlim, type='l', log='x', ylab=sprintf("%sWrapped Phase, rad", ifelse(can.unwrap,'Un-',''))) -->
<!--   abline(h=-pi, lty=2) -->
<!--   plot(f, G, type='l', xlim=xlim, log='xy', ylab="Gain", xlab="Frequency, Hz") -->
<!-- }) -->

We can safely assume that the spectral density estimates for
periods longer than $\approx 100$ seconds will be either
spurious, or lacking in seismic energy, so we will exclude them.
```{r, echo=TRUE}
csd <- data.frame(f, p, lf, Coh, k, G, G.err, Phi = Phi * 180 / pi)
csd.f <- subset(csd, p <= 100)
```
We see that the phase and gain are quite stable across most of the seismic band,
but coherence degrades at frequencies above $\approx$ 0.1 Hz.
Accordingly, the uncertainty grows very large as the coherence goes
to zero at very high frequency.
The reason for the loss in coherence is related to the hydraulic
diffusivity of aquifer system and the details of the well sampling 
pressure in it.
```{r, echo=FALSE, fig.show='hold', fig.width=6., fig.height=6.5}

layout(matrix(1:3), heights=c(2,2,1.5))
par(oma=c(1,1,3,1), cex=0.8, las=1, tcl=-0.2, mgp=c(2,0.3,0))

par(mar=c(0.1, 4, 1, 4))
plot(Coh ~ f, csd.f, 
     log='x',
     xlab='', ylab='',
     type='l', xaxt='n', 
     ylim=c(-0.6,1), 
     xaxs='i', yaxt='n',
     yaxs='i', frame=FALSE)
abline(h=c(0,1), lty=3)
mtext('Coherence', font=2, line=-2.5, adj=0.3)
mtext('No. tapers', side=1, font=3, line=-2.7, adj=0.2)
axis(2, at=seq(0,1,by=0.2))
#axis(1, col=NA, col.ticks=1, labels=FALSE)
axis(3, col=NA, col.ticks=1)
title("Cross Spectrum: Pressure from Strain (Tohoku M9)", outer=TRUE)
par(new=TRUE)
plot(K, xaxs='i', yaxs='i', axes=FALSE, ylim=c(0,1000),xlab='', ylab='')
axis(4, at=seq(0,300,by=100))
par(new=FALSE)
box()

par(mar=c(0.1, 4, 0, 4))
nsig <- 2
with(csd.f, {
  lG <- log10(G)
  Delt <- nsig*log10(exp(1))*G.err/G
  Upper <- lG + Delt
  Lower <- lG - Delt
  ylim <- range(c(lG)) + log10(2)*c(-1,2)
  plot(f, lG, type='l', 
       log='x',
       yaxt='n', xlab='', ylab='',
       xaxt='n', 
       col=NA, frame=FALSE,
       xaxs='i', yaxs='i', ylim=ylim)
  polygon(c(f, rev(f)), c(Upper, rev(Lower)), col='lightcyan', border='lightgrey')
  lmsc <- log10(abs(strain_scaling))
  abline(h=lmsc, col=2, lty=2)
  mtext("scaling\nfrom lm", side=4, at=lmsc, col=2, line=0.2, font=3)
  lines(f, lG)
})
mtext('Admittance', font=2, line=-4.3, adj=0.3)
mtext(parse(text=sprintf("%s * sigma ~ 'uncert.'", nsig)), cex=1, adj=0.3, line=-5.5,  col='cyan4')
ll <- c(1,2,5)
lbls <- c(ll/1000, ll/100, ll/10, ll, ll*10)
ats <- log10(lbls)
lbls[c(F,T,T)] <- ""
axis(2, at=ats, labels=lbls)
box()

par(mar=c(1, 4, 0, 4))
degadd <- 180
plot(Phi + degadd ~ f, csd.f, 
     log='x',
     type='l', #col='lightgrey', 
     xlab='Frequency, Hz', ylab="Degrees (<0 = lag)",
     xaxs='i', frame=FALSE,
     #yaxs='i', 
     yaxt='n')

lmphs <- 180 * sign(strain_scaling) + degadd
abline(h=lmphs, col=2, lty=2)
mtext("sign of\nlm coef.", side=4, at=lmphs, col=2, line=0.2, font=3)

mtext(sprintf('Relative Phase',ifelse(can.unwrap," (Unwrapped)","")), font=2, line=-4.0, adj=0.3)
axis(2)
axis(1, at=10**(-3:1), labels=paste0("(",c(1000,100,10,1,"1/10"),'s)'), line=1.6, lwd=0)
box()
```

All of this functionality is available in the function `cross_spectrum`.
<!-- In comparison with -->
<!-- a Welch-type CSD -- calculated by setting `k=NULL`, the sine multitaper result is -->
<!-- more accurate across the full frequency band, and does not degrade at low frequencies: -->
<!-- ```{r, fig.width=6., fig.height=3.5} -->
<!-- TohCS <- cross_spectrum(toh.dat, k=50, verbose=FALSE) -->
<!-- TohCS_welch <- cross_spectrum(toh.dat, k=NULL, verbose=FALSE) # turn off k to get a Welch overlapping csd -->
<!-- plot(Admittance ~ Period, TohCS, col=NA, log='x', main="Pore Pressure from Strain: Tohoku", xlab="Period, sec") -->
<!-- lines(Admittance ~ Period, TohCS_welch, col='salmon') -->
<!-- lines(Admittance ~ Period, TohCS, lwd=2) -->
<!-- ``` -->


## References

Barbour, A. J., (2015), Pore-Pressure Sensitivities to Dynamic Strains: Observations in Active Tectonic Regions, 
Journal of Geophysical Research: Solid Earth, 120, 5863 — 5883,
[DOI: 10.1002/2015JB012201](https://doi.org/10.1002/2015JB012201)

Priestley, M. B. (1981), Spectral analysis and time series,
Academic Press, New York
